---
title: "Converting {qs} to {qs2} with rig"
author: Benjamin L. Hlina
date: "2026-09-25"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{using_qs_and_qs2_with_rig}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, knitr-settings, include = FALSE}
knitr::opts_chunk$set(
  eval = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction 

This vignette will walk through how to easily convert old `{qs}` file format to `{qs2}` file format using the R Installation Manager ([rig](https://rig.r-lib.org/)) to install R v4.3 which allows the installation and use of both `{qs}` and `{qs2}`. This will first involve installing rig which requires writing some commands in terminal. I do suggest for this conversion using [Positron](https://positron.posit.co/) instead of [RStudio](https://posit.co/products/open-source/rstudio) as doing this work will be more comfortable using Positron. However, if getting Positron installed and set up feels difficult and change is hard everything below can be done in RStudio. 

# Installing and setting up rig

Installation instructions for ([rig](https://rig.r-lib.org/)) depend on the operating system of the computer you are 
working on and can be found in the [installation guide](https://rig.r-lib.org/install.html) for rig. You will need 
to select both the operating system and whether to install rig for your home directory or system-wide. Please 
follow this guide as this installation is quite straightforward with commands needing to be run in terminal. 

Once installed we can use rig again in terminal in either admin or user roles. Currently, user role is newer and may be a bit
more prone to bugs. 

To use rig in admin mode please follow [admin mode tutorial](https://rig.r-lib.org/tutorial-admin.html). From this tutorial we 
can install R v4.3 which we will need to open `.qs` files and convert them to `.qs2` files. 

Because admin mode installs as administrative user in all operating systems, rig needs administrator rights, and asks for your computer's password. 

First, we can see what version(s) of R are currently installed using the following command: 

```{bash, rig-list}
rig list
```

Next we can see what versions of R are available in rig using
```{bash, rig-available}
rig available 
``` 
We can then install R v4.3 by using the following: 

```{bash, rig-install-r-4.3}
rig add 4.3
```

At this point we could switch the default R version to 4.3 if this was something we desired and can be obtained using: 

```{bash, default-r}
rig default 4.3
```

However, you likely have other versions of R (e.g., 4.6) that you are currently using. So, it is probably best to leave this setting alone. 
 
# Installing {qs} and {qs2} 

## Positron 

We can now open a new session in Positron. If we go to the top right portion of the console, we will see a plus sign appear with a down arrow next to it. This down arrow, when clicked, will allow us to start a new R session using different versions. When clicking the down arrow, you will see the current version of R and `Start Another...` as options. Click `Start Another...` which will open a dropdown showing different versions of R to start a session. At this point select R 4.3. This will start a new console using R 4.3. 

Once started we can move to install the packages. 

## RStudio

To use RStudio we need to use the following commands in terminal. First let's look at what versions of R are installed. 

```{bash, rstudio-available-r}
rig list
```

The operating system and version of RStudio are important here. On macOS arm64 computers (non-Intel) rig prefers arm64 R, unless an exact version match is only available with x86_64 R. For Windows machines, `rig rstudio` needs RStudio Desktop 2021.09.0+351 or later.

Next, we can launch RStudio using the following for macOS (non-Intel) computers

```{bash, launch-rstudio-with-r4.3-macOS}
rig rstudio 4.3-arm64
```

On Windows machines we can use the following: 

```{bash, launch-rstudio-with-r4.3-windows}
rig rstudio 4.3
```

Once started we can move to install the packages. 

## Install packages 
Next, we can install `{qs}` and `{qs2}` using the following commands and follow the prompts: 

```{r, install-qs-qs2}
pak::pkg_install(c(
  "url::https://cran.r-project.org/src/contrib/Archive/qs/qs_0.27.3.tar.gz",
  "qs2"
))
```

At this point we can open `.qs` files using the following as an example: 

```{r, open-qs} 
df <- qs::qread(file = 'path_to_qs_file.qs')
```

Next, we can save the file as a `.qs2` file using `{qs2}`

```{r, save-qs2}
qs2::qs_save(object = df, file = 'path_to_save_qs2_file.qs2')
```

Congratulations! At this point we have successfully converted old `.qs` files to `.qs2` files and can now use them in the most up-to-date version of R and `{qs2}`.  
