---
title: "qs2"
output:
  html_vignette:
    keep_md: no
  rmarkdown::github_document: default
vignette: >
  %\VignetteIndexEntry{qs2}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, setup, echo=FALSE}
IS_GITHUB <- Sys.getenv("IS_GITHUB") != ""
```

```{r results='asis', echo=FALSE, eval=IS_GITHUB}
cat('
[![R-CMD-check](https://github.com/traversc/qs2/workflows/R-CMD-check/badge.svg)](https://github.com/traversc/qs2/actions)
[![CRAN-Status-Badge](http://www.r-pkg.org/badges/version/qs2)](https://cran.r-project.org/package=qs2)
[![CRAN-Downloads-Badge](https://cranlogs.r-pkg.org/badges/qs2)](https://cran.r-project.org/package=qs2)
[![CRAN-Downloads-Total-Badge](https://cranlogs.r-pkg.org/badges/grand-total/qs2)](https://cran.r-project.org/package=qs2)
')
```

*qs2: a framework for efficient serialization*

`qs2` is the successor to the `qs` package. The goal is to have relaible and fast performance for saving and loading objects in R.

The `qs2` format directly uses R serialization (via the `R_Serialize`/`R_Unserialize` C API) while improving compression and writing/reading to disk. 
If you are familiar with the `qs` package, the benefits and usage are the same.

```{r eval=FALSE}
qs_save(data, "myfile.qs2")
data <- qs_read("myfile.qs2")
```

Use the file extension `qs2` to distinguish it from the original `qs` package. It is not compatible with the original `qs` format.


# Installation

```{r eval=FALSE}
install.packages("qs2")
```

On Mac or Linux, Tt enable multithreading (using `Intel Thread Building Blocks` via the `RcppParallel` package) compile from source. It is enabled by default on Windows. 

```{r eval=FALSE}
remotes::install_cran("qs2", type = "source", configure.args = " --with-TBB --with-simd=AVX2")
```


# Advanced usage: converting qs2 to RDS

Because the `qs2` format directly uses R serialization you can convert it to RDS and vice versa.

```{r eval=FALSE}
file_qs2 <- tempfile(fileext = ".qs2")
file_rds <- tempfile(fileext = ".RDS")
x <- runif(1e6)

# save `x` with qs_save
qs_save(x, file_qs2)

# convert the file to RDS
qs_to_rds(input_file = file_qs2, output_file = file_rds)

# read `x` back in with `readRDS`
xrds <- readRDS(file_rds)
stopifnot(identical(x, xrds))
```

# Advanced usage: the qdata format

The package also introduces the `qdata` format which has its own serialization layout and works with only data types (vectors, lists, data frames, matrices). 

It will replace internal types (functions, promises, external pointers, environments, objects) with NULL. The `qdata` format differs from 
the `qs2` format in that it is NOT a general.

If you have clean data and understand the caveats, you may want to use it. The eventual goal of `qdata` is to also have interoperability with other languages, particularly `Python`. 

```{r eval=FALSE}
qd_save(data, "myfile.qs2")
data <- qd_read("myfile.qs2")
```

## Benchmarks

A summary table across 4 datasets is presented below.

| Algorithm       | nthreads | Save Time | Read Time | Compression |
|-----------------|----------|-----------|-----------|-------------|
| qs2             | 1        | 4.41      | 3.32      | 2.56        |
| qdata           | 1        | 4.08      | 3.11      | 2.58        |
| qs              | 1        | 4.24      | 3.17      | 2.51        |
| saveRDS         | 1        | 27.21     | 6.84      | 2.32        |
| base::serialize | 1        | 10.17     | 11.51     | 1.10        |
| fst             | 1        | 2.76      | 3.80      | 1.59        |
| parquet         | 1        | 4.99      | 3.47      | 2.33        |
| qs2             | 4        | 1.36      | 2.73      | 2.56        |
| qdata           | 4        | 1.28      | 2.51      | 2.58        |
| qs              | 4        | 1.84      | 2.64      | 2.51        |
| fst             | 4        | 2.66      | 3.76      | 1.59        |
| parquet         | 4        | 4.93      | 3.08      | 2.33        |

**Notes on running each algorithm**

* `qs2`, `qdata` and `qs` used `compress_level = 5`
* `parquet` (from the `arrow` package) were run with `compression = zstd` and `compress_level = 5`
* `fst` used `compress_level = 55`
* `base::serialize` was run with `ascii = FALSE` and `xdr = FALSE`

**Datasets**

* `enwik8` the first 1E8 lines of English Wikipedia
* `gaia` galactic coordinates and color of 7.2 million stars
* `tcell` the genomic sequencing of a COVID patient's immune system

These datasets are openly licensed and represent a combination of numeric and textual data across multiple domains. 

See `inst/benchmarks` on Github for comprehensive comparisons, reproducibility and data sources.


